---
title: "RAPID Alpha Diversity Group Significance"
author: "Becca Maher"
date: "6/6/2019"
output: html_document
---
## Script for calculating alpha diversity statistics and testing for significant differences with metadata variables

### 1 Setup the working space
```{r libraries, include=FALSE}
rm(list=ls())
library(phyloseq)
library(multcomp)
library(picante)
library(ggplot2 )
library(tidyverse)
library(FSA)
library(MASS)
```

#### Functions
```{r functions}
# Arc-sine transformation
asinTransform <- function(p) { asin(sqrt(p)) }
# Calculate standard error
sderr <- function(x) {sd(x)/sqrt(length(x))}
#+++++++++++++++++++++++++
# Function to calculate the mean and the standard deviation
# for each group
#+++++++++++++++++++++++++
# data : a data frame
# varname : the name of a column containing the variable
#to be summariezed
# groupnames : vector of column names to be used as
# grouping variables
data_summary <- function(data, varname, groupnames){
  require(plyr)
  summary_func <- function(x, col){
    c(mean = mean(x[[col]], na.rm=TRUE),
      sd = sderr(x[[col]]), na.rm=TRUE)
  }
  data_sum<-ddply(data, groupnames, .fun=summary_func,
                  varname)
  data_sum <- rename(data_sum, c("mean" = varname))
  return(data_sum)
}
```

#### Load the data
Load the alphadiv.csv files created in the RAPID_alpha_statsA.Rmd script. First starting with all species
```{r load}
alphadiv <- read.csv("~/Box Sync/RAPID/RAPID-analysis/data/alphadiv.csv")
head(alphadiv)
```


### 2 Data Exploration
Looking for normality and homogeneity of group dispersions
#### Richness
```{r}
# box plot of richness by species
ggplot(alphadiv, aes(x = species, y = richness)) +  geom_boxplot()
# box plot of richness by time
ggplot(alphadiv, aes(x = time, y = richness)) +  geom_boxplot()
# box plot of richness by treatment
ggplot(alphadiv, aes(x = treatment, y = richness)) +  geom_boxplot()
# outlier is acropora, T3, Nitrate
# Testing for normality
histogram(alphadiv$richness)
qqnorm(alphadiv$richness)
abline(0,1)
shapiro.test(alphadiv$richness)
histogram(log(alphadiv$richness))
qqnorm(log(alphadiv$richness))
abline(0,1)
shapiro.test(log(alphadiv$richness))
histogram(sqrt(alphadiv$richness))
qqnorm(sqrt(alphadiv$richness))
abline(0,1)
shapiro.test(sqrt(alphadiv$richness))
# test for homogeneity of group dispersions
bartlett.test(sqrt(richness) ~ time, data = alphadiv)
bartlett.test(sqrt(richness) ~ treatment, data = alphadiv)
bartlett.test(sqrt(richness) ~ species, data = alphadiv)
# richness data are normal with equal group variances with a sqrt transformation
```
#### Evenness
```{r}
# box plot of evenness by species
ggplot(alphadiv, aes(x = species, y = evenness)) +  geom_boxplot()
# box plot of evenness by time
ggplot(alphadiv, aes(x = time, y = evenness)) +  geom_boxplot()
# box plot of evenness by treatment
ggplot(alphadiv, aes(x = treatment, y = evenness)) +  geom_boxplot()
# outlier is acropora, T3, Nitrate
# Testing for normality
histogram(alphadiv$evenness)
qqnorm(alphadiv$evenness)
abline(0,1)
shapiro.test(alphadiv$evenness) # non-normal
histogram(asinTransform(alphadiv$evenness))
qqnorm(asinTransform(alphadiv$evenness))
abline(0,1)
shapiro.test(asinTransform(alphadiv$evenness)) # non-normal
# test for homogeneity of group dispersions
bartlett.test(evenness ~ time, data = alphadiv)
bartlett.test(evenness ~ treatment, data = alphadiv)
bartlett.test(evenness ~ species, data = alphadiv)
# AsinTransformation does not make the evenness data normal
# Variances are equal for groups under un-transformed data
# will proceed with a non-Parametric Kruskal-Wallis test for evenness data
```
#### Faith's phylogenetic diversity
```{r}
# box plot of faithPD by species
ggplot(alphadiv, aes(x = species, y = faithPD)) +  geom_boxplot()
# box plot of faithPD by time
ggplot(alphadiv, aes(x = time, y = faithPD)) +  geom_boxplot()
# box plot of faithPD by treatment
ggplot(alphadiv, aes(x = treatment, y = faithPD)) +  geom_boxplot()
# outlier is acropora, T3, Nitrate
# Testing for normality
histogram(alphadiv$faithPD)
qqnorm(alphadiv$faithPD)
abline(0,1)
shapiro.test(alphadiv$faithPD) # non-normal
histogram(sqrt(alphadiv$faithPD))
qqnorm(sqrt(alphadiv$faithPD))
abline(0,1)
shapiro.test(sqrt(alphadiv$faithPD)) # non-normal
# test for homogeneity of group dispersions
bartlett.test(faithPD ~ time, data = alphadiv)
bartlett.test(faithPD ~ treatment, data = alphadiv)
bartlett.test(faithPD ~ species, data = alphadiv)

bartlett.test(sqrt(faithPD) ~ time, data = alphadiv)
bartlett.test(sqrt(faithPD) ~ treatment, data = alphadiv)
bartlett.test(sqrt(faithPD) ~ species, data = alphadiv) # sig
# Log transformation does not make the faithPD data normal, but sqrt-transform does
# Variances are equal for groups under un-transformed data and sqrt-transformed data
# Non-Parametric Kruskal-Wallis test for untransformed or ANOVA for transformed?
```

### 3 Analysis
#### Chao1 data (richness)
Proceeding with sqrt-transformed richness data
```{r rich}
summary(aov(sqrt(richness) ~ time, data = alphadiv))
summary(aov(sqrt(richness) ~ treatment, data = alphadiv)) # not sig
summary(aov(sqrt(richness) ~ species, data = alphadiv))

TukeyHSD(aov(sqrt(richness) ~ time, data = alphadiv))
# TukeyHSD(aov(sqrt(richness) ~ treatment, data = alphadiv))
TukeyHSD(aov(sqrt(richness) ~ species, data = alphadiv))
```

#### Simpson's index data (evenness)
Proceeding with non-parametric kruskal wallis test from Simpsons data
```{r even}
kruskal.test(evenness ~ time, data = alphadiv)
kruskal.test(evenness ~ treatment, data = alphadiv) # not sig
kruskal.test(evenness ~ species, data = alphadiv) # not sig

dunnTest(evenness ~ time, data = alphadiv, method = "bh")
```
#### Faith's phylogenetic diversity data (faithPD)
Proceeding with sqrt-transformed faithPD data
```{r faithPD}
summary(aov(sqrt(faithPD) ~ time, data = alphadiv))
summary(aov(sqrt(faithPD) ~ treatment, data = alphadiv)) # not sig
summary(aov(sqrt(faithPD) ~ species, data = alphadiv))

TukeyHSD(aov(sqrt(faithPD) ~ time, data = alphadiv))
# TukeyHSD(aov(sqrt(faithPD) ~ treatment, data = alphadiv))
TukeyHSD(aov(sqrt(faithPD) ~ species, data = alphadiv))
```
Check for interactions
```{r int}
summary(aov(sqrt(richness) ~ time*treatment*species, data = alphadiv))
summary(aov(sqrt(faithPD) ~ time*treatment*species, data = alphadiv))
# This interaction test with Kruskal wallis is not comparable to anova
int <- interaction(alphadiv$time, alphadiv$treatment, alphadiv$species)
kruskal.test(evenness ~ int, data = alphadiv)
int_sig <- dunnTest(evenness ~ int, data = alphadiv, method = "bh")
intdf <- as.data.frame(int_sig$res) 
intdf <- as.data.frame(int_sig$res)
new <- intdf[order(intdf$P.adj),]
head(new) # only three significant interactions
```

### For Acropora only
Subsetting the data to only include Acropora corals
```{r}
alphadiv <- alphadiv[which(alphadiv$species =="ACR"),]
summary(alphadiv)
```
#### Chao1 data (richness)
Proceeding with sqrt-transformed richness data
```{r rich2}
summary(aov(sqrt(richness) ~ time, data = alphadiv))
summary(aov(sqrt(richness) ~ treatment, data = alphadiv)) # not sig

TukeyHSD(aov(sqrt(richness) ~ time, data = alphadiv))
# TukeyHSD(aov(sqrt(richness) ~ treatment, data = alphadiv))
```

#### Simpson's index data (evenness)
Proceeding with non-parametric kruskal-wallis test for simpson's data
```{r even2}
kruskal.test(evenness ~ time, data = alphadiv)
kruskal.test(evenness ~ treatment, data = alphadiv) # not sig

dunnTest(evenness ~ time, data = alphadiv, method = "bh")
```
#### Faith's phylogenetic diversity data (faithPD)
Proceeding with sqrt-transformed faithPD data
```{r faithPD2}
summary(aov(sqrt(faithPD) ~ time, data = alphadiv))
summary(aov(sqrt(faithPD) ~ treatment, data = alphadiv)) # not sig

TukeyHSD(aov(sqrt(faithPD) ~ time, data = alphadiv))
# TukeyHSD(aov(sqrt(faithPD) ~ treatment, data = alphadiv))
```
Check for interactions
```{r int2}
summary(aov(sqrt(richness) ~ time*treatment, data = alphadiv))
summary(aov(sqrt(faithPD) ~ time*treatment, data = alphadiv))
# This interaction test with Kruskal wallis is not comparable to anova
int <- interaction(alphadiv$time, alphadiv$treatment)
kruskal.test(evenness ~ int, data = alphadiv)
int_sig <- dunnTest(evenness ~ int, data = alphadiv, method = "bh")
intdf <- as.data.frame(int_sig$res)
new <- intdf[order(intdf$P.adj),]
head(new) # no sig
```