---
title: "RAPID_beta_stats"
author: "Becca Maher"
date: "August 12, 2019"
output: html_document
---

## 1 Set up
```{r, include=FALSE}
## clear workspace------------------------
rm(list=ls())

# load libraries
library('phyloseq'); packageVersion('phyloseq')
library('vegan'); packageVersion('vegan')
library("usedist")


## functions----------------------
pairwise.adonis.dm <- function(x,factors,stratum=NULL,p.adjust.m="bonferroni",perm=999){
  
  library(vegan)
  
  if(class(x) != "dist") stop("x must be a dissimilarity matrix (dist object)")
  
  co = as.matrix(combn(unique(factors),2))
  
  pairs = c()
  
  F.Model =c()
  
  R2 = c()
  
  p.value = c()
  
  
  
  for(elem in 1:ncol(co)){
    
    sub_inds <- factors %in% c(as.character(co[1,elem]),as.character(co[2,elem]))
    
    resp <- as.matrix(x)[sub_inds,sub_inds]
    
    ad = adonis(as.dist(resp) ~
                  
                  factors[sub_inds], strata=stratum[sub_inds], permutations=perm);
    
    pairs = c(pairs,paste(co[1,elem],'vs',co[2,elem]));
    
    F.Model =c(F.Model,ad$aov.tab[1,4]);
    
    R2 = c(R2,ad$aov.tab[1,5]);
    
    p.value = c(p.value,ad$aov.tab[1,6])
    
  }
  
  p.adjusted = p.adjust(p.value,method=p.adjust.m)
  
  pairw.res = data.frame(pairs,F.Model,R2,p.value,p.adjusted)
  
  return(pairw.res)
  }
```

```{r}
# First import qd object, this is the phyloseq object from the import_qiime_to_phyloseq.R
load(file = "~/Box Sync/RAPID/RAPID-analysis/data/qd_881.RData")
qd
```

## 2 Analysis of log transformed data

Optional transformations to the data
```{r}
# Log-transform OTU table
otus_log <- as(otu_table(qd), "matrix")
otus_log <- decostand(otus_log, method = "log")
OTU_log <- otu_table(otus_log, taxa_are_rows = TRUE)
otu_table(qd) <- OTU_log
```

### Calculate distance matrices

```{r}
# Calculate all four distance matrices: weighted unifrac, unweighted unifrac, bray curtis, binary jaccard
qd_wu <- phyloseq::distance(qd, method = "wunifrac")
qd_un <- phyloseq::distance(qd, method = "unifrac")
qd_bc <- phyloseq::distance(qd, method = "bray")
qd_bj <- distance(qd, method = "jaccard", binary =TRUE)
```
### PERMANOVAs
```{r}
# PERMANOVA's with Adonis -  Permutational Multivariate Analasis of Variance
# Make a data frame from the sample_data
sampledf <- data.frame(sample_data(qd))
```

#### Interaction
```{r}
# Adonis test for between group diversity, with full formula
adonis(qd_bc ~ time*species*treatment, data = sampledf) # sig: time, species, time:species
adonis(qd_bj ~ time*species*treatment, data = sampledf) # sig: time, species, time:species
adonis(qd_wu ~ time*species*treatment, data = sampledf) # sig: time, species
adonis(qd_un ~ time*species*treatment, data = sampledf) # sig: time, species, time:species
```
Pairwise comparisons
```{r}
# all significant
pairwise.adonis.dm(qd_bc, sample_data(qd)$time, p.adjust.m = "fdr")
pairwise.adonis.dm(qd_bj, sample_data(qd)$time, p.adjust.m = "fdr")
pairwise.adonis.dm(qd_wu, sample_data(qd)$time, p.adjust.m = "fdr")
pairwise.adonis.dm(qd_un, sample_data(qd)$time, p.adjust.m = "fdr")
# All significant
pairwise.adonis.dm(qd_bc, sampledf$species, p.adjust.m = "fdr")
pairwise.adonis.dm(qd_bj, sampledf$species, p.adjust.m = "fdr")
pairwise.adonis.dm(qd_wu, sampledf$species, p.adjust.m = "fdr")
pairwise.adonis.dm(qd_un, sampledf$species, p.adjust.m = "fdr")
```

### PERMDISPs
#### Species
```{r}
# PERMDISP with betadisper - Multivariate Homogeneity of group dispersions
anova(betadisper(qd_bc, sampledf$species, bias.adjust = TRUE)) # non sig
anova(betadisper(qd_bj, sampledf$species, bias.adjust = TRUE)) # non sig
anova(betadisper(qd_wu, sampledf$species, bias.adjust = TRUE)) # significant
anova(betadisper(qd_un, sampledf$species, bias.adjust = TRUE)) # non sig
# pairwise test
p.adjust(permutest(betadisper(qd_wu, sampledf$species, 
                              bias.adjust = TRUE), 
                   pairwise=TRUE)$pairwise$permuted, method = 'fdr')
```
#### Treatment
```{r}
anova(betadisper(qd_bc, sampledf$treatment, bias.adjust = TRUE))
anova(betadisper(qd_bj, sampledf$treatment, bias.adjust = TRUE))
anova(betadisper(qd_wu, sampledf$treatment, bias.adjust = TRUE))
anova(betadisper(qd_un, sampledf$treatment, bias.adjust = TRUE))
# All non-significant
```
#### Time
```{r}
anova(betadisper(qd_bc, sampledf$time, bias.adjust = TRUE)) # non sig
anova(betadisper(qd_bj, sampledf$time, bias.adjust = TRUE)) # significant
anova(betadisper(qd_wu, sampledf$time, bias.adjust = TRUE)) # significant
anova(betadisper(qd_un, sampledf$time, bias.adjust = TRUE)) # significant
# pairwise tests
p.adjust(permutest(betadisper(qd_bj, sampledf$time, 
                              bias.adjust = TRUE), 
                   pairwise=TRUE)$pairwise$permuted, method = 'fdr')
p.adjust(permutest(betadisper(qd_wu, sampledf$time, 
                              bias.adjust = TRUE), 
                   pairwise=TRUE)$pairwise$permuted, method = 'fdr')
p.adjust(permutest(betadisper(qd_un, sampledf$time, 
                              bias.adjust = TRUE), 
                   pairwise=TRUE)$pairwise$permuted, method = 'fdr')

```
#### Interaction
```{r}
anova(betadisper(qd_bc, sampledf$time.trt, bias.adjust = TRUE))
anova(betadisper(qd_bj, sampledf$time.trt, bias.adjust = TRUE))
anova(betadisper(qd_un, sampledf$time.trt, bias.adjust = TRUE)) # significant
anova(betadisper(qd_wu, sampledf$time.trt, bias.adjust = TRUE)) # significant
# pairwise tests
p.adjust(permutest(betadisper(qd_un, sampledf$time.trt, 
                              bias.adjust = TRUE), 
                   pairwise=TRUE)$pairwise$permuted, method = 'fdr')
p.adjust(permutest(betadisper(qd_wu, sampledf$time.trt, 
                              bias.adjust = TRUE), 
                   pairwise=TRUE)$pairwise$permuted, method = 'fdr')

```

## 3 Analysis of log transformed Acropora data
```{r}
# subset by species
qd_acr <- subset_samples(qd, species =="ACR")
qd_acr
```

Calculate distance matrices
```{r}
# Calculate all four distance matrices: weighted unifrac, unweighted unifrac, bray curtis, binary jaccard
qd_acr_wu <- phyloseq::distance(qd_acr, method = "wunifrac")
qd_acr_un <- phyloseq::distance(qd_acr, method = "unifrac")
qd_acr_bc <- phyloseq::distance(qd_acr, method = "bray")
qd_acr_bj <- distance(qd_acr, method = "jaccard", binary =TRUE)
```

### PERMANOVAs
```{r }
# PERMANOVA's with Adonis -  Permutational Multivariate Analasis of Variance
# Make a data frame from the sample_data
sampledf <- data.frame(sample_data(qd_acr))
```
#### Interaction
```{r}
adonis(qd_acr_bc ~ time*treatment, data = sampledf) # time
adonis(qd_acr_bj ~ time*treatment, data = sampledf) # time
adonis(qd_acr_wu ~ time*treatment, data = sampledf) # time
adonis(qd_acr_un ~ time*treatment, data = sampledf) # time
```
#### Pairwise
```{r}
# pairwise tests
pairwise.adonis.dm(qd_acr_bc, sample_data(qd_acr)$time, p.adjust.m = "fdr")
pairwise.adonis.dm(qd_acr_bj, sample_data(qd_acr)$time, p.adjust.m = "fdr")
pairwise.adonis.dm(qd_acr_wu, sample_data(qd_acr)$time, p.adjust.m = "fdr")
pairwise.adonis.dm(qd_acr_un, sample_data(qd_acr)$time, p.adjust.m = "fdr")
```


### PERMDISPs
#### Treatment
```{r}
# PERMDISP with betadisper - Multivariate Homogeneity of group dispersions
anova(betadisper(qd_acr_bc, sampledf$treatment, bias.adjust = TRUE)) # significant
anova(betadisper(qd_acr_bj, sampledf$treatment, bias.adjust = TRUE))
anova(betadisper(qd_acr_wu, sampledf$treatment, bias.adjust = TRUE))
anova(betadisper(qd_acr_un, sampledf$treatment, bias.adjust = TRUE))
# pairwise
p.adjust(permutest(betadisper(qd_acr_wu, sampledf$treatment, 
                              bias.adjust = TRUE), 
                   pairwise=TRUE)$pairwise$permuted, method = 'fdr')
```
#### Time
```{r}
anova(betadisper(qd_acr_bc, sampledf$time, bias.adjust = TRUE))
anova(betadisper(qd_acr_bj, sampledf$time, bias.adjust = TRUE)) # sig
anova(betadisper(qd_acr_wu, sampledf$time, bias.adjust = TRUE)) # sig
anova(betadisper(qd_acr_un, sampledf$time, bias.adjust = TRUE))
# pairwise
p.adjust(permutest(betadisper(qd_acr_bj, sampledf$time, 
                              bias.adjust = TRUE), 
                   pairwise=TRUE)$pairwise$permuted, method = 'fdr')
p.adjust(permutest(betadisper(qd_acr_wu, sampledf$time, 
                              bias.adjust = TRUE), 
                   pairwise=TRUE)$pairwise$permuted, method = 'fdr')
```

## 4 Analysis of log transformed Pocillopora data
```{r}
# subset by species
qd_poc <- subset_samples(qd, species =="POC")
qd_poc
```

Calculate distance matrices
```{r}
# Calculate all four distance matrices: weighted unifrac, unweighted unifrac, bray curtis, binary jaccard
qd_poc_wu <- phyloseq::distance(qd_poc, method = "wunifrac")
qd_poc_un <- phyloseq::distance(qd_poc, method = "unifrac")
qd_poc_bc <- phyloseq::distance(qd_poc, method = "bray")
qd_poc_bj <- distance(qd_poc, method = "jaccard", binary =TRUE)
```

### PERMANOVAs
```{r}
# PERMANOVA's with Adonis -  Permutational Multivariate Analasis of Variance
# Make a data frame from the sample_data
sampledf <- data.frame(sample_data(qd_poc))
```
#### Interaction
```{r}
adonis(qd_poc_bc ~ time*treatment, data = sampledf) # time
adonis(qd_poc_bj ~ time*treatment, data = sampledf) # time
adonis(qd_poc_wu ~ time*treatment, data = sampledf) # time
adonis(qd_poc_un ~ time*treatment, data = sampledf) # time
```
#### Pairwise
```{r}
# pairwise tests
pairwise.adonis.dm(qd_poc_bc, sample_data(qd_poc)$time, p.adjust.m = "fdr")
pairwise.adonis.dm(qd_poc_bj, sample_data(qd_poc)$time, p.adjust.m = "fdr")
pairwise.adonis.dm(qd_poc_wu, sample_data(qd_poc)$time, p.adjust.m = "fdr")
pairwise.adonis.dm(qd_poc_un, sample_data(qd_poc)$time, p.adjust.m = "fdr")
```

### PERMDISPs
#### Time
```{r}
anova(betadisper(qd_poc_bc, sampledf$time, bias.adjust = TRUE))
anova(betadisper(qd_poc_bj, sampledf$time, bias.adjust = TRUE)) # sig
anova(betadisper(qd_poc_wu, sampledf$time, bias.adjust = TRUE)) # sig
anova(betadisper(qd_poc_un, sampledf$time, bias.adjust = TRUE))
# pairwise
p.adjust(permutest(betadisper(qd_poc_bj, sampledf$time, 
                              bias.adjust = TRUE), 
                   pairwise=TRUE)$pairwise$permuted, method = 'fdr')
p.adjust(permutest(betadisper(qd_poc_wu, sampledf$time, 
                              bias.adjust = TRUE), 
                   pairwise=TRUE)$pairwise$permuted, method = 'fdr')
## 5 Analysis of log transformed Porites data
```{r}
# subset by species
qd_por <- subset_samples(qd, species =="POR")
qd_por
```

Calculate distance matrices
```{r}
# Calculate all four distance matrices: weighted unifrac, unweighted unifrac, bray curtis, binary jaccard
qd_por_wu <- phyloseq::distance(qd_por, method = "wunifrac")
qd_por_un <- phyloseq::distance(qd_por, method = "unifrac")
qd_por_bc <- phyloseq::distance(qd_por, method = "bray")
qd_por_bj <- distance(qd_por, method = "jaccard", binary =TRUE)
```

### PERMANOVAs
```{r}
# PERMANOVA's with Adonis -  Permutational Multivariate Analasis of Variance
# Make a data frame from the sample_data
sampledf <- data.frame(sample_data(qd_por))
```
#### Interaction
```{r}
adonis(qd_por_bc ~ time*treatment, data = sampledf) # time
adonis(qd_por_bj ~ time*treatment, data = sampledf) # time
adonis(qd_por_wu ~ time*treatment, data = sampledf) # time
adonis(qd_por_un ~ time*treatment, data = sampledf) # time
```
#### Pairwise
```{r}
# pairwise tests
pairwise.adonis.dm(qd_por_bc, sample_data(qd_por)$time, p.adjust.m = "fdr")
pairwise.adonis.dm(qd_por_bj, sample_data(qd_por)$time, p.adjust.m = "fdr")
pairwise.adonis.dm(qd_por_wu, sample_data(qd_por)$time, p.adjust.m = "fdr")
pairwise.adonis.dm(qd_por_un, sample_data(qd_por)$time, p.adjust.m = "fdr")
```

### PERMDISPs
#### Time
```{r}
anova(betadisper(qd_por_bc, sampledf$time, bias.adjust = TRUE))
anova(betadisper(qd_por_bj, sampledf$time, bias.adjust = TRUE)) # sig
anova(betadisper(qd_por_wu, sampledf$time, bias.adjust = TRUE)) # sig
anova(betadisper(qd_por_un, sampledf$time, bias.adjust = TRUE))
# pairwise
p.adjust(permutest(betadisper(qd_por_bj, sampledf$time, 
                              bias.adjust = TRUE), 
                   pairwise=TRUE)$pairwise$permuted, method = 'fdr')
p.adjust(permutest(betadisper(qd_por_wu, sampledf$time, 
                              bias.adjust = TRUE), 
                   pairwise=TRUE)$pairwise$permuted, method = 'fdr')