---
title: "RAPID Calculate Alpha Diversity Statistics"
author: "Becca Maher"
date: "6/6/2019"
output: html_document
---
## Script for calculating alpha diversity statistics and testing for significant differences with metadata variables

### 1 Setup the working space
#### Libraries
```{r libraries, message=FALSE}
rm(list=ls())
library(phyloseq)
library(multcomp)
library(picante)
library(ggplot2 )
library(tidyverse)
```

#### Functions
```{r functions, include = FALSE}
# Arc-sine transformation
asinTransform <- function(p) { asin(sqrt(p)) }
# Estimate faith's phylogenetic diversity 
estimate_pd <- function(phylo){
  
  # Error if input is not of class phylo
  if(class(phylo) != "phyloseq"){
    stop("Input file is not of class 'phyloseq'.")
  }
  
  # Error if no class phy_tree
  if(!(.hasSlot(phylo, "phy_tree"))){
    stop("Could not find tree slot in phylo object.")
  }
  
  # Transpose if needed
  # Adapted from phyloseq/vegan import
  OTU <- phyloseq::otu_table(phylo)
  if (taxa_are_rows(OTU)) {
    OTU <- t(OTU)
  }
  
  # Get matrix version of OTU table
  otutable <- as(OTU, "matrix")
  
  # Get phylogenetic tree from pyloseq object
  tree <- phyloseq::phy_tree(phylo)
  
  # Print status message
  message("Calculating Faiths PD-index...")
  
  # If object is greater than 10mb, then print status message
  if(object.size(otutable) > 10000000){
    message("This is a large object, it may take awhile...")
  }
  
  # Calculate Faith's PD-index
  pdtable <- picante::pd(otutable, tree, include.root = F)
  
  # Return data frame of results
  return(pdtable)
}
# Calculate standard error
sderr <- function(x) {sd(x)/sqrt(length(x))}
#+++++++++++++++++++++++++
# Function to calculate the mean and the standard deviation
# for each group
#+++++++++++++++++++++++++
# data : a data frame
# varname : the name of a column containing the variable
#to be summariezed
# groupnames : vector of column names to be used as
# grouping variables
data_summary <- function(data, varname, groupnames){
  require(plyr)
  summary_func <- function(x, col){
    c(mean = mean(x[[col]], na.rm=TRUE),
      sd = sderr(x[[col]]), na.rm=TRUE)
  }
  data_sum<-ddply(data, groupnames, .fun=summary_func,
                  varname)
  data_sum <- rename(data_sum, c("mean" = varname))
  return(data_sum)
}
```

#### Load the data
For all alpha and beta diversity statistics, I will be using the rarefied sOTU table. The table has been rarefied to 1778 sequences per sample.
```{r load}
load("~/Box Sync/RAPID/RAPID-analysis/data/qd_rar.RData")
qd
```

### Calculating alpha diversity statistics
Initialize matrices to store richness and evenness estimates
```{r init}
nsamp = nsamples(qd)

richness <- matrix(nrow = nsamp)
row.names(richness) <- sample_names(qd)

evenness <- matrix(nrow =nsamp)
row.names(evenness) <- sample_names(qd)

faithPD <- matrix(nrow = nsamp)
row.names(faithPD) <- sample_names(qd)
##
```
Calculate statistics
```{r stats, message = FALSE}
# Options for measures = ("Observed", "Chao1", "ACE", "Shannon", "Simpson", "InvSimpson", "Fisher")

# Calculate richness
rich <- as.numeric(as.matrix(subset(estimate_richness(qd, measures = "Chao1"), select = c(1))))
richness[ ,] <- rich
colnames(richness) [1] <- "richness"

# Calculate evenness
even <- as.numeric(as.matrix(estimate_richness(qd, measures = "Simpson")))
evenness[ ,] <- even
colnames(evenness) [1] <- "evenness"

# Calculate Faith's PD
faith <- as.numeric(as.matrix(subset(estimate_pd(qd), select = c(1))))  # estimate_pd is a function assigned at the end of the script
faithPD[ ,] <- faith
colnames(faithPD) [1] <- "faithPD"

# Included the subset in "rich" and "faith" because the Chao1 and Faith's PD measurement outputs two measures per sample (Chao1 and se.chao1)
# and we only want Chao1, so we select for the first column
```
Combine our estimates for richness, evenness, and faith's PD into one dataframe
```{r compile}
alpha <- cbind(richness, evenness[,1][match(rownames(richness), rownames(evenness))],
               faithPD[,1][match(rownames(richness), rownames(faithPD))])
colnames(alpha) [2] <- "evenness"
colnames(alpha) [3] <- "faithPD"
head(alpha)

# Add the sample metadata into this dataframe
# Definitely not the cleanest way to do this...
colnames(sample_data(qd))[1] <- "SampleID"
s <- data.frame(sample_data(qd))
rownames(s) <- NULL
alpha <- cbind(SampleID = rownames(alpha), alpha)
rownames(alpha) <- NULL
alphadiv <- cbind(alpha, s)
alphadiv <- alphadiv[,-5]
alphadiv$richness <- as.numeric(as.character(alphadiv$richness))
alphadiv$evenness <- as.numeric(as.character(alphadiv$evenness))
alphadiv$faithPD <- as.numeric(as.character(alphadiv$faithPD))
head(alphadiv)
#write.csv(alphadiv, file = "~/Box Sync/RAPID/RAPID-analysis/data/alphadiv.csv")
```